---
title: "predPLSR"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    logo: logo.png
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(MASS)
library(caret)
library(tidyverse)
library(dplyr)
library(DT)
library(shiny)
library(shinydashboard)
library(Hmisc)
library(shinyWidgets)
library(GGally)
library(patchwork)

```


Summary {data-orientation=rows}
======================================================================

Column {.sidebar data.width=300 .box }
-------------

*Data - Upload*


```{r}

options(shiny.maxRequestSize=30*1024^2)

      fileInput("file1", "Choose Train CSV File",
                      multiple = FALSE,
                      accept = c("text/csv",
                               "text/comma-separated-values,text/plain",
                               ".csv"))
      
      fileInput("file2", "Choose Test CSV File",
                      multiple = FALSE,
                      accept = c("text/csv",
                               "text/comma-separated-values,text/plain",
                               ".csv"))
      
      dfTrain <- reactive({
        req(input$file1)
        read_csv(input$file1$datapath)
      })
      
      dfTest <- reactive({
        req(input$file2)
        read_csv(input$file2$datapath)
      })


```

Row {data-height=100}
-----------------------------------------------------------------------

### Trainset Size

```{r}
renderValueBox({
trainSize <- nrow(dfTrain())
valueBox(trainSize, " ")
})
```


### Testset Size

```{r}
renderValueBox({
testSize <- nrow(dfTest())
valueBox(testSize, " ")
})
```

### No of Replicates
```{r}
renderValueBox({
no.rep <- length(unique(dfTrain()$rep))
valueBox(no.rep, " ")
})
```

### Dependent Variable
```{r}
renderValueBox({
  valueBox("Melamine Concentration" , " ")
})
```

### Independent Variables

```{r}
renderValueBox({
  valueBox("Absorbances correspond to wave numbers", " ")
})
```

Row 
--------------------------------------------------------

### Trainset melamine concentration composition

```{r}

renderPlot({

  melamine <- unique(dfTrain()$mel.concentration)
  animalMix <- 100 - melamine
  concentraion <- melamine
  
  data <- data.frame(concentraion, melamine, animalMix)
  
  data <- data %>%
    pivot_longer(2:3, names_to = "element", values_to = "value") %>%
    arrange(desc(concentraion)) 
  
  data <- data[data$element == "melamine",]
  
  data <- data[data$value != 0,]
  
  data <- data[data$value != 100,]
  
  data$concentraion <- paste0(rep("con",times=nrow(data)),1:nrow(data))
  
  data1 <- data %>% arrange(value) %>%  
    mutate(concentraion = concentraion %>% as_factor %>% fct_inorder) 
  
  ggplot(data1, aes(fill=element, y=value, x=concentraion)) + 
    geom_col() +
    coord_flip()+
    ylab("Melamine content (%)")+
    xlab("")+
    theme(legend.position = "") +
    scale_fill_manual(values = "#0d98ba")
})

```


### Testset melamine concentration composition

```{r}
renderPlot({
  melamine <- unique(dfTest()$mel.concentration)
  animalMix <- 100 - melamine
  concentraion <- melamine
  
  data <- data.frame(concentraion, melamine, animalMix)
  
  data <- data %>%
    pivot_longer(2:3, names_to = "element", values_to = "value") %>%
    arrange(desc(concentraion)) 
  
  data <- data[data$element == "melamine",]
  
  data <- data[data$value != c(0, 100),]
  
  data$concentraion <- paste0(rep("con",times=nrow(data)),1:nrow(data))
  
  data1 <- data %>% arrange(value) %>%  
    mutate(concentraion = concentraion %>% as_factor %>% fct_inorder) 
  
  ggplot(data1, aes(fill=element, y=value, x=concentraion)) + 
    geom_col() +
    coord_flip()+
    ylab("Melamine content (%)")+
    xlab("")+
    theme(legend.position = "")+
    scale_fill_manual(values = "pink")
})


```


Spectrum plots
=========

Column {.sidebar data.width=300 .box }
-------------

```{r}
renderUI({
  df_new <- dfTrain()[dfTrain()$mel.concentration != 0, ]
  df_new <- df_new[df_new$mel.concentration != 100, ]
selectInput("select_con","Choose Concentration", choices = sort(unique(df_new$mel.concentration)))
})
```

Column {.tabset}
-------------------------------------
   
### Pure Melamine

```{r}

renderPlot({
  melamine_con1 <- dfTrain()[dfTrain()$mel.concentration == 100, ]
  melamine_con1 <- melamine_con1[, c(-1, -2)]
  col.num <- ncol(melamine_con1)
  melamine_con1 <- melamine_con1 %>% pivot_longer(2:col.num, names_to = "wave_num", values_to = "abs" )
  melamine_con1$wave_num <- as.numeric(melamine_con1$wave_num)
  ggplot(data = melamine_con1, aes(x=wave_num, y=abs, col=rep)) +
    geom_line() +
    xlab(bquote("Wave Number " (cm ^ -1))) +
    ylab("Absorbance") +
    ggtitle("Pure Melamine Mix") +
    theme(legend.position = "bottom") +
    labs(color = "Replicate")
  
})

```   
 
### Pure Animal Mix
    
```{r}
renderPlot({
  melamine_con1 <- dfTrain()[dfTrain()$mel.concentration == 0, ]
  melamine_con1 <- melamine_con1[, c(-1, -2)]
  col.num <- ncol(melamine_con1)
  melamine_con1 <- melamine_con1 %>% pivot_longer(2:col.num, names_to = "wave_num", values_to = "abs" )
  melamine_con1$wave_num <- as.numeric(melamine_con1$wave_num)
  ggplot(data = melamine_con1, aes(x=wave_num, y=abs, col=rep)) +
    geom_line() +
    xlab(bquote("Wave Number " (cm ^ -1))) +
    ylab("Absorbance") +
    ggtitle("Pure Animal Mix") +
    theme(legend.position = "bottom") +
    labs(color = "Replicate")
})

```

### Melamine concentration on choice
    
```{r}

renderPlot({
  melamine_con1 <- dfTrain()[dfTrain()$mel.concentration == input$select_con, ]
  melamine_con1 <- melamine_con1[, c(-1, -2)]
  col.num <- ncol(melamine_con1)
  melamine_con1 <- melamine_con1 %>% pivot_longer(2:col.num, names_to = "wave_num", values_to = "abs" )
  melamine_con1$wave_num <- as.numeric(melamine_con1$wave_num)
  ggplot(data = melamine_con1, aes(x=wave_num, y=abs, col=rep)) +
    geom_line() +
    xlab(bquote("Wave Number " (cm ^ -1))) +
    ylab("Absorbance") +
    theme(legend.position = "bottom") +
    labs(color = "Replicate")
})

```

Model Details {data-orientation=columns}
=====

Column {.sidebar data.width=300 .box }
-------------


```{r}

      sliderInput("peakRange", "Select the wave number range",  min = 450, max=5500, value = c(3400,3500), step = 100)
      
      
      minRange <- reactive({
        req(input$peakRange[1])
      })
      
       maxRange <- reactive({
        req(input$peakRange[2])
      })


```


Column {data-width=450}
------

### Model Description

```{r}

renderPrint({
  
  trainData <- dfTrain()[dfTrain()$mel.concentration != 100.00, ]
  trainData1 <- trainData[, c(-1,-3)]
  trainData2 <- trainData1[, c(1, (minRange()-448):(maxRange()-448))]

  testData1 <- dfTest()[, c(-1,-3)]
  testData2 <- testData1[, c(1, (minRange()-448):(maxRange()-448))]
  
  test.features <- testData2[, c(-1)]
  concentration.test <- testData2[, c(1)]
  
  pls.n <- length(unique(dfTrain()$mel.concentration)) + 5
  
  set.seed(1)

ctrl <- trainControl(
  method = "LOOCV",
  number = 10,
)



model <- train(
  mel.concentration ~ .,
  data = trainData2,
  method = 'pls',
  trControl = ctrl,
  tuneLength = pls.n
)

# Predictions
pls_pred.test <- predict(model, newdata = test.features)

model

})

```


column {data-width=500}
------

### Scree plot

```{r}
renderPlot({
  
  trainData <- dfTrain()[dfTrain()$mel.concentration != 100.00, ]
  trainData1 <- trainData[, c(-1,-3)]
  trainData2 <- trainData1[, c(1, (minRange()-448):(maxRange()-448))]

  testData1 <- dfTest()[, c(-1,-3)]
  testData2 <- testData1[, c(1, (minRange()-448):(maxRange()-448))]
  
  test.features <- testData2[, c(-1)]
  concentration.test <- testData2[, c(1)]
  
  pls.n <- length(unique(dfTrain()$mel.concentration)) + 5
  
  set.seed(1)

ctrl <- trainControl(
  method = "LOOCV",
  number = 10,
)



model <- train(
  mel.concentration ~ .,
  data = trainData2,
  method = 'pls',
  trControl = ctrl,
  tuneLength = pls.n
)

# Predictions
pls_pred.test <- predict(model, newdata = test.features)
plot(model)

})
```

Predictions {data-orientation=columns}
=====

Column {.sidebar data.width=300 .box }
-------------


```{r}

      sliderInput("peakRange", "Select the wave number range",  min = 450, max=5500, value = c(3400,3500), step = 100)
      
      
      minRange <- reactive({
        req(input$peakRange[1])
      })
      
       maxRange <- reactive({
        req(input$peakRange[2])
      })


```

Column {data-width=450}
------

### RMSE

```{r}
renderValueBox({
  
  trainData <- dfTrain()[dfTrain()$mel.concentration != 100.00, ]
  trainData1 <- trainData[, c(-1,-3)]
  trainData2 <- trainData1[, c(1, (minRange()-448):(maxRange()-448))]

  testData1 <- dfTest()[, c(-1,-3)]
  testData2 <- testData1[, c(1, (minRange()-448):(maxRange()-448))]
  
  test.features <- testData2[, c(-1)]
  concentration.test <- testData2[, c(1)]
  
  pls.n <- length(unique(dfTrain()$mel.concentration)) + 5
  
  set.seed(1)

ctrl <- trainControl(
  method = "LOOCV",
  number = 10,
)



model <- train(
  mel.concentration ~ .,
  data = trainData2,
  method = 'pls',
  trControl = ctrl,
  tuneLength = pls.n
)

# Predictions
pls_pred.test <- predict(model, newdata = test.features)

# calculate RMSE
rmse <- round(sqrt(mean((pls_pred.test - concentration.test[,1,drop=TRUE])^2)), digits = 3)
valueBox(rmse, "RMSE")

})

```

### Actual Vs. Predicted
  
```{r}
renderPlot({
  
  trainData <- dfTrain()[dfTrain()$mel.concentration != 100.00, ]
  trainData1 <- trainData[, c(-1,-3)]
  trainData2 <- trainData1[, c(1, (minRange()-448):(maxRange()-448))]

  testData1 <- dfTest()[, c(-1,-3)]
  testData2 <- testData1[, c(1, (minRange()-448):(maxRange()-448))]
  
  test.features <- testData2[, c(-1)]
  concentration.test <- testData2[, c(1)]
  
  pls.n <- length(unique(dfTrain()$mel.concentration)) + 5
  
  set.seed(1)

ctrl <- trainControl(
  method = "LOOCV",
  number = 10,
)



model <- train(
  mel.concentration ~ .,
  data = trainData2,
  method = 'pls',
  trControl = ctrl,
  tuneLength = pls.n
)

# Predictions
pls_pred.test <- predict(model, newdata = test.features)

df <- data.frame(actual = concentration.test[,1,drop=TRUE], predicted = pls_pred.test)

# Plot actual vs predictions
ggplot(df,aes(actual, predicted)) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x, color = "red") +
  labs(x = "Actual Concentration", y = "Predicted Concentration")
})


```


column {data-width=400}
------

### R squared

```{r}
renderGauge({
  
  trainData <- dfTrain()[dfTrain()$mel.concentration != 100.00, ]
  trainData1 <- trainData[, c(-1,-3)]
  trainData2 <- trainData1[, c(1, (minRange()-448):(maxRange()-448))]

  testData1 <- dfTest()[, c(-1,-3)]
  testData2 <- testData1[, c(1, (minRange()-448):(maxRange()-448))]
  
  test.features <- testData2[, c(-1)]
  concentration.test <- testData2[, c(1)]
  
  pls.n <- length(unique(dfTrain()$mel.concentration)) + 5
  
  set.seed(1)

ctrl <- trainControl(
  method = "LOOCV",
  number = 10,
)

model <- train(
  mel.concentration ~ .,
  data = trainData2,
  method = 'pls',
  trControl = ctrl,
  tuneLength = pls.n
)

# Predictions
pls_pred.test <- predict(model, newdata = test.features)

rsq <- cor(concentration.test, pls_pred.test) ^ 2
gauge(rsq*100, min = 0, max = 100,
      gaugeSectors(success = c(70,100),
                   warning = c(40,69),
                   danger = c(0,39)))

})


renderText({
  
  trainData <- dfTrain()[dfTrain()$mel.concentration != 100.00, ]
  trainData1 <- trainData[, c(-1,-3)]
  trainData2 <- trainData1[, c(1, (minRange()-448):(maxRange()-448))]

  testData1 <- dfTest()[, c(-1,-3)]
  testData2 <- testData1[, c(1, (minRange()-448):(maxRange()-448))]
  
  test.features <- testData2[, c(-1)]
  concentration.test <- testData2[, c(1)]
  
  pls.n <- length(unique(dfTrain()$mel.concentration)) + 5
  
  set.seed(1)

ctrl <- trainControl(
  method = "LOOCV",
  number = 10,
)

model <- train(
  mel.concentration ~ .,
  data = trainData2,
  method = 'pls',
  trControl = ctrl,
  tuneLength = pls.n
)

# Predictions
pls_pred.test <- predict(model, newdata = test.features)

rsq <- cor(concentration.test, pls_pred.test) ^ 2

paste("Model explians", round(rsq*100, digits=3),"% of the total variability of the Melamine concentration.")
})
```

### Download Actual and Predicted Concentrations

```{r}

renderDataTable({
  
  trainData <- dfTrain()[dfTrain()$mel.concentration != 100.00, ]
  trainData1 <- trainData[, c(-1,-3)]
  trainData2 <- trainData1[, c(1, (minRange()-448):(maxRange()-448))]

  testData1 <- dfTest()[, c(-1,-3)]
  testData2 <- testData1[, c(1, (minRange()-448):(maxRange()-448))]
  
  test.features <- testData2[, c(-1)]
  concentration.test <- testData2[, c(1)]
  
  pls.n <- length(unique(dfTrain()$mel.concentration)) + 5
  
  set.seed(1)

ctrl <- trainControl(
  method = "LOOCV",
  number = 10,
)



model <- train(
  mel.concentration ~ .,
  data = trainData2,
  method = 'pls',
  trControl = ctrl,
  tuneLength = pls.n
)

# Predictions
pls_pred.test <- predict(model, newdata = test.features)

df <- data.frame('actual' = concentration.test[,1,drop=TRUE], 'predicted' = pls_pred.test)

datatable(
  df, extensions ='Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)

})

```

About {data-orientation=columns}
======================================================================
This dashboard was created to ease out the PLSR analysis for spectrum data. Dashboard is built with R using the R Markdown framework.


**Created by**

Kavya Atapattu, Thejaka Hewakuruppu, Savindu Kodikara, Madhusha Perera, Udeshi Salgado, Anjana Senadeera, Department of Statistics, University of Sri Jayewardenepura

[Thiyanga S. Talagala](https://thiyanga.netlify.app/), Department of Statistics, University of Sri Jayewardenepura (Project supervisor)

**Code**

The code behind the dashboard is available on [GitHub](https://github.com/StatConG2/predPLSR).


Coulumn {data-height=800}
-----------------------------------------------------------------------


**Data**

To proceed with the analysis data should be in the the following format given in Figure 01.

```{r, fig.cap="Figure 01: Data format", fig.align = 'center'}
knitr::include_graphics ("data.jpg")

```


**Citation**

Talagala T., Atapattu A. M. K. N., Hewakuruppu D. T., Kodikara S. V., Perera M. M. I., Salgado M. U. P., Senadeera B. S. A. (2023). predPLSR. Statistical Consultancy Services, Department of Statistics, USJ. https://github.com/StatConG2/predPLSR